---
title: "Employee Attrition Analysis for IBM HR Data"
---

```{r, echo = FALSE}
library(tidyverse)    
library(magrittr)    
library(cluster)    
library(cluster.datasets)    
library(cowplot)    
library(NbClust)    
library(clValid)    
library(ggfortify)    
library(clustree)    
library(dendextend)    
library(factoextra)    
library(FactoMineR)    
library(corrplot)    
library(GGally)    
library(ggiraphExtra)    
library(knitr)    
library(kableExtra)
library(caret)
library(alluvial)
library(ROCR)

```


```{r}
#load data saved as rds
hrDataRaw <- read_rds("IBM-HR-Data-Full.rds")
hrDataRaw <- as.data.frame(hrDataRaw)
glimpse(hrDataRaw)
```


## Preprocessing Data
```{r}
#remove variables that are not useful
row.names(hrDataRaw) <- hrDataRaw$EmployeeNumber
hrData <- hrDataRaw%>%select(-StandardHours, -Over18, -EmployeeCount, -EmployeeNumber)

# convert attrition to binary
hrData$AttrBin <- if_else(hrData$Attrition=="Yes",1,0)

# convert overtime to binary
hrData$OverTimeBin <- if_else(hrData$OverTime == "Yes", 1, 0)
```


```{r}
#check for missing values
sum(!complete.cases(hrData))

#in detail
sapply(hrData, function(x) sum(is.na(x)))
```
There are no missing values, no need to impute/fill in values.


```{r}
#convert categorical variables to factors
hrData$Attrition<-factor(hrData$Attrition)
hrData$BusinessTravel<-factor(hrData$BusinessTravel)
hrData$Department<-factor(hrData$Department)
hrData$Education <- factor(hrData$Education)
hrData$EducationField<-factor(hrData$EducationField)
hrData$EnvironmentSatisfaction <- factor(hrData$EnvironmentSatisfaction)
hrData$Gender<-factor(hrData$Gender)
hrData$JobRole<-factor(hrData$JobRole)
hrData$JobInvolvement <- factor(hrData$JobInvolvement)
hrData$JobLevel <- factor(hrData$JobLevel)
hrData$JobSatisfaction <- factor(hrData$JobSatisfaction)
hrData$MaritalStatus<-factor(hrData$MaritalStatus)
hrData$OverTime<-factor(hrData$OverTime)
hrData$PerformanceRating <- factor(hrData$PerformanceRating)
hrData$RelationshipSatisfaction <- factor(hrData$RelationshipSatisfaction)
hrData$StockOptionLevel <- factor(hrData$StockOptionLevel)
hrData$WorkLifeBalance <- factor(hrData$WorkLifeBalance)
```


# Data Exploration
```{r}
#distribution of the dependent variable
table(hrData$Attrition)
```

As expected, the data is not distributed evenly. We will need to further understand the data to decide how we can go about it.

## Understanding the data
```{r}
#break down the data into two different dataset
# Attrition= Yes
nonAttr <- hrData %>% filter(Attrition=="Yes")

#Attrition = No
attr <- hrData %>% filter(Attrition=="No")

table(nonAttr$JobRole)
table(attr$JobRole)

table(nonAttr$BusinessTravel)
table(attr$BusinessTravel)


#for the continuous variables, lets check the mean
mean(nonAttr$MonthlyIncome)
mean(attr$MonthlyIncome)

mean(nonAttr$PercentSalaryHike)
mean(attr$PercentSalaryHike)

mean(nonAttr$YearsSinceLastPromotion)
mean(attr$YearsSinceLastPromotion)

mean(nonAttr$Age)
mean(attr$Age)

```

```{r}
boxplot(
  Age~Attrition,
  data=hrData,
  main="Age vs Attrition",
  xlab="Attrition",
  ylab="Age")
```

```{r}
boxplot(MonthlyIncome~Attrition,data=hrData, main="Monthly Income vs Attrition", 
   xlab="Attrition", ylab="Income per month")
```


## Finding patterns in data
```{r}
p <- hrData %>%
  group_by(Gender, BusinessTravel, Attrition)%>%
  count()
  
alluvial(p[,1:2], freq=p$n, border=NA,
         #hide = p$n < quantile(p$n,.50),
         col=ifelse(p$Attrition == "Yes", "red", "gray"))

female <- hrData %>% filter(Gender == "Female")
femaleFrequentTraveller <- hrData %>% filter(Gender == "Female", BusinessTravel == "Travel_Frequently")

table(female$Attrition)
table(femaleFrequentTraveller$Attrition)
```
The frequent travelling female seem to be leaving the company at higher rate than the regular females.



### More Exploration
```{r}
#create new object with all the categorical variables removed
hrDataNum <- hrData %>% select(AttrBin, Age, DailyRate, DistanceFromHome, HourlyRate, MonthlyIncome, MonthlyRate, NumCompaniesWorked, OverTimeBin, PercentSalaryHike, TotalWorkingYears, TrainingTimesLastYear, YearsAtCompany, YearsInCurrentRole, YearsSinceLastPromotion, YearsWithCurrManager)

#view table
summary(hrDataNum) %>% kable() %>% kable_styling()
```


What’s the relationship between the different attributes? 
Use `corrplot()` to create correlation matrix.
```{r}
#correlation plot
corrplot(cor(hrDataNum), type = "full", method = "ellipse", tl.cex = 0.9)
```


```{r}
# also create an object with only categorical variables
hrDataCat <- hrData%>%select(Attrition, BusinessTravel, Department, Education, EducationField, EnvironmentSatisfaction, Gender, JobRole, JobInvolvement, JobLevel, JobSatisfaction, MaritalStatus, OverTime, PerformanceRating, RelationshipSatisfaction, StockOptionLevel, WorkLifeBalance)
summary(hrDataCat) %>% kable() %>% kable_styling()
```


# Build Model
1. Create a base model: with only numerical variables
2. Evaluate the model
3. Fix issues
4. Optimization

In all we have 34 features consisting of both the categorical as well as the numerical features. The target variable is the 'Attrition' of the employee which can be either a Yes or a No. This is what we have to predict.

Hence this is a Binary Classification problem.

```{r}
# Build train & test
set.seed(9876)
train <- sample(1:nrow(hrDataNum), nrow(hrDataNum)*.7)
test <- -train

trainData <- hrDataNum[train,]
testData <- hrDataNum[test,]
```


```{r}
#start with a full model
model1 <- glm(AttrBin~., data=trainData, family = binomial(link="logit"))

summary(model1)
```
The overtime indicator, age, distance from home, number of companies worked are the most significant indicators in this model.


```{r}
#looking for multi collinearity
car::vif(model1)
```
There is problems with multi-collinearity with the total working years, monthly income, years at company, years in current role which could be removed while building the next model.


```{r}
# Accesing the predective ability of the logistic regression model
predictTest1 <- predict(model1, newdata = testData, type='response')

#setting up cutoff at 0.5
predictTest1 <- if_else((predictTest1 >= 0.5), 1, 0)

# Create confusion matrix
caret::confusionMatrix(factor(predictTest1), factor(testData$AttrBin))
```
Gives an accuracy of 85% and a sensitivity of 96%, but the multi collinearity issue is significant. We need to build a better model.


```{r}
# Plotting the ROC curve
predictTrain1 <- predict(model1, trainData, type = "response")
ROCRPred <- prediction(predictTrain1, trainData$AttrBin)
ROCRPerf <- performance(ROCRPred,"tpr","fpr")
plot(ROCRPerf,colorize = TRUE, print.cutoffs.at = seq(0.1, by = 0.1))
```

