---
title: "Employee Attrition Analysis for IBM HR Data"
---

```{r, echo=FALSE}
library(readxl)
library(tidyverse)
library(moments)
library(car)
library(lmerTest)
library(pROC)
```


```{r}
#load data saved as rds
hrData <- read_rds("IBM-HR-Data-Full.rds")

#view variables
str(hrData)
```


## Preprocessing Data
```{r}
#check for missing values
sum(!complete.cases(hrData))
```
There are no missing values, no need to impute/fill in values.


```{r}
#convert categorical variables to factors
hrData$Attrition<-factor(hrData$Attrition)
hrData$BusinessTravel<-factor(hrData$BusinessTravel)
hrData$Department<-factor(hrData$Department)
hrData$EducationField<-factor(hrData$EducationField)
hrData$Gender<-factor(hrData$Gender)
hrData$JobRole<-factor(hrData$JobRole)
hrData$MaritalStatus<-factor(hrData$MaritalStatus)
hrData$Over18<-factor(hrData$Over18)
hrData$OverTime<-factor(hrData$OverTime)

#remove variables that are not useful for analysis
hrData <- hrData%>%select(-EmployeeCount,-EmployeeNumber)

#taking a glimpse of how the data looks
glimpse(hrData)
```


# Data Exploration
```{r}
#distribution of the dependent variable
table(hrData$Attrition)
```

As expected, the data is not distributed evenly. We will need to further understand the data to decide how we can go about it.

## Understanding the data
```{r}
#break down the data into two different dataset
# Attrition= Yes
nonAttr <- hrData %>% filter(Attrition=="Yes")

#Attrition = No
attr <- hrData %>% filter(Attrition=="No")



table(nonAttr$JobRole)
table(attr$JobRole)

table(nonAttr$BusinessTravel)
table(attr$BusinessTravel)


#for the continuous variables, lets check the mean
mean(nonAttr$MonthlyIncome)
mean(attr$MonthlyIncome)

mean(nonAttr$WorkLifeBalance)
mean(attr$WorkLifeBalance)

mean(nonAttr$PercentSalaryHike)
mean(attr$PercentSalaryHike)

mean(nonAttr$YearsSinceLastPromotion)
mean(attr$YearsSinceLastPromotion)

mean(nonAttr$StandardHours)
mean(attr$StandardHours)

mean(nonAttr$PerformanceRating)
mean(attr$PerformanceRating)

mean(nonAttr$RelationshipSatisfaction)
mean(attr$RelationshipSatisfaction)


```

```{r}
#check distribution of age & other categorical variables
table(nonAttr = nonAttr$Age)
table(attr = attr$Age)


mean(nonAttr$Age)
mean(attr$Age)


boxplot(
  Age~Attrition,
  data=hrData,
  main="Age vs Attrition",
  xlab="Attrition",
  ylab="Age")
```

```{r}
boxplot(MonthlyIncome~Attrition,data=hrData, main="Monthly Income vs Attrition", 
   xlab="Attrition", ylab="Income per month")
```


```{r}
plot(MonthlyIncome~Attrition, data= hrData)
```


## Finding patterns in data
```{r}
female <- hrData %>% filter(Gender == "Female")
femaleFrequentTraveller <- hrData %>% filter(Gender == "Female", BusinessTravel == "Travel_Frequently")



table(female$Attrition)
table(femaleFrequentTraveller$Attrition)
```
The frequent travelling female seem to be leaving the company at higher rate than the regular females.


```{r}
input_ones <- hrData[which(hrData$Attrition == "Yes"), ]  # all 1's
input_TWO <- hrData[which(hrData$Attrition == "No"), ]  # all 0's
set.seed(100)  # for repeatability of samples
input_ones_training_rows <- sample(1:nrow(input_ones), 0.7*nrow(input_ones))  # 1's for training
input_twos_training_rows <- sample(1:nrow(input_TWO), 0.7*nrow(input_ones))  # 0's for training. Pick as many 0's as 1's
training_ones <- input_ones[input_ones_training_rows, ]  
training_Two <- input_TWO[input_twos_training_rows, ]
trainingData <- rbind(training_ones, training_Two)  # row bind the 1's and 0's \

table(trainingData$Attrition)
```

##Creating Test Data
```{r}

str(hrData)
testones<- input_ones[-input_ones_training_rows,]
testTwo<- input_TWO[-input_twos_training_rows,]

testData <- rbind(testones, testTwo)

```

##CREATING A MODEL

```{r}
colnames(hrData)
## converting varibales into factor
hrData$BusinessTravel<- as.factor(hrData$BusinessTravel)
hrData$EducationField<-as.factor(hrData$EducationField)
hrData$Gender<- as.factor(hrData$Gender)
hrData$JobRole<- as.factor(hrData$JobRole)
str(hrData)

model<- glm(Attrition~., data=df, family = binomial)
summary(model)
car::vif(model)
Predicted<- predict(model, trainingData, type ="response")
View(Predicted)

Predictedtest<-predict(model, df2, type ="response")

length(unique(hrData$StandardHours))

```


#
```{r}
roc(df2$Attrition,Predictedtest)     #roc for testdata
#plot ROC
plot.roc(testData$Attrition, Predictedtest)   #roc plot for testdata

```

#creating model with only few variables.

```{r}
str(hrData)
model <- glm(Attrition ~ Age + JobRole + TrainingTimesLastYear + YearsSinceLastPromotion + WorkLifeBalance, data=trainingData, family = binomial)
summary(model)

Predicted<- predict(model, trainingData, type ="response")
View(Predicted)

Predictedtest<-predict(model, testData, type ="response")

roc(testData$Attrition,Predictedtest)     #roc for testdata

#plot ROC
plot.roc(testData$Attrition, Predictedtest)   #roc plot for testdata
```


#Creating Graphs

```{r}
plot(hrData$Attrition, hrData$Age, xlab = "ATTRITION", ylab = "AGE",col="red")

plot(hrData$Attrition, hrData$BusinessTravel, xlab = "ATTRITION", ylab = "Business Travels",col="Yellow")

plot(hrData$Attrition, hrData$Gender, xlab = "ATTRITION", ylab = "Gender",col="blue")

plot(hrData$Attrition, hrData$JobRole, xlab = "ATTRITION", ylab = "JobRole",col="Yellow")

plot(hrData$Attrition, hrData$TrainingTimesLastYear, xlab = "ATTRITION", ylab = "Business Travels",col="Yellow")
```

```{r}
M<-cor(hrData)
head(round(M,2))
corrplot(corr, method="circle")
```

```{r}
head(hrData)
```

```{r}
unique(hrData$BusinessTravel)
Correlation_check = hrData%>%mutate_if(is.factor,as.numeric)
corrplot(cor(Correlation_check),method = "number")

M<-cor(hrData)
head(round(M,2))
corrplot(M, method="number")
```

```{r}
#check multi-collinearity of the model
car::vif(model)
```

